In-Band Synchronization for\\Distributed SDN Control Planes

### 2015-06-01 inband-sync.tex
phrasing

\begin{abstract}
Control planes of forthcoming Software-Defined Networks (SDNs)
will be \emph{distributed}: to ensure availability and fault-tolerance,
to improve load-balancing, and to reduce overheads,
control modules operating on the data plane should be physically distributed.
%a redundant control plane
%can ensure a high availability, and a spatially distributed
%control plane allows to handle time-critical dataplane
%events close to their origins.
%[[PK this is a contraversial point, not important for us
%Moreover, an SDN may be managed by multiple administrators simultaneously.
%]]
However, %a distributed control plane also introduces
%new challenges,
in order to guarantee \emph{consistency} of network operation, actions
performed on the data plane by different controllers may need to be \emph{synchronized},  which is a
 nontrivial task.
%[[PK sounds contradicting to the fact that such systems already exist
%However, not much is known today about how to implement
%such a distributed control plane.
%]]
%
In this paper, we propose a synchronization
framework for control planes  based on the well-known concept of
\emph{transactions}.
%
A transaction aggregates a sequence
of control messages and specific synchronization primitives,
providing all-or-nothing semantics.
We argue that a powerful transactional interface should and can be
implemented \emph{in-band}, using
standard OpenFlow commands on the data plane configuration.
%Indeed, recent OpenFlow versions already come with useful but primitive
%features, in particular \emph{bundling} and \emph{rule-overlap checks}.
%In particular, we present a
%transactional interface which supports arbitrary atomic
%transactions
%over the switch configuration space:
%Our transactions
%can include complex notions of conflicts (beyond flow overlap checks),
%as well as computations and multiple controller interactions.
We discuss how to use our synchronization framework to implement
 consistent policy composition and fault-tolerant control-plane applications.
\end{abstract}

@@

We believe that our synchronization mechanisms
can be useful in many situations.
For example, the pathological scenario described in
Figure~\ref{fig:CAS-example} can be resolved easily if we put
the policy update in our transactional $\execatomic$ construct together
 with a \emph{Compare-and-Set} (CAS) operation on the policy identifier.
A $\cas(\addr,\textit{old},\textit{new})$ operation on a memory location $\addr$
checks if the content of $\addr$ is $\textit{old}$ and if so,
replaces it with $\textit{new}$ and returns $\true$ (the CAS
\emph{succeeds}), otherwise it returns $\false$ (the CAS
\emph{fails} which causes an abort of
the invoking transaction).
It is straightforward to implement the CAS operation by executing
$\textit{\compare}(\addr, old)$ followed by $\textit{\memwrite}(\addr,
new)$ within the {\execatomic} construct.

@@

More generally, CAS can be used to solve the fundamental
problem of fault-tolerant \emph{consensus}~\cite{FLP85}
\hide{%[[PK, feels like we do not really need this level of detail here
 in which the controllers
need to agree on one of their private inputs, despite failures of an
arbitrary number of controllers.
Indeed, each controller simply performs, within a transaction, a CAS operation to replace the default value
in a memory location at the meta-configuration of a data plane switch.
If the transaction commits, the controller considers itself
the \emph{winner} and decides on its input,
if it aborts (the controller fails in the CAS), it remains to read the
memory to get the input of the winner and decide on it.
}
and allows us
to implement any generic replicated state machine~\cite{Her91}
in the control plane,  in the \emph{wait-free} manner, i.e., tolerating
asynchrony and failure of any number of controllers.
As such, CAS can also be used to implement STN~\cite{stn},
by providing the postulated \emph{read-modify-write} object.

@@

Our mechanism can also be used as a generic template to
implement concurrent policy compositions of a user-specific
$\textsc{\ufunc}()$ function~\cite{stn}, e.g., our load-balancing
function $\textsc{lb\_update}$ (Example~1 in Section~\ref{sec:motivation}).
That is, let us consider a simple policy update protocol that
allows a set of  controllers to concurrently update switch policies, i.e., sets of
%[[PK policies have been defined in Background]]
effective flow entries, under the
condition that these rules can be \emph{composed} with the currently installed
policy.
Algorithm~\ref{alg:simple-update} uses the bundle feature to update
the switch configuration, where the configuration contains both the
policy and a meta-configuration memory address $\paddr$ that holds the \emph{policy identifier (pid)} .
In the algorithm, the controller first reads the currently installed
policy together with its {\pid},  applies the
$\textsc{\ufunc}()$ function to the current policy which results in a
set of update FlowMod commands and then tries to apply them
atomically together with a CAS operation on $\paddr$ replacing $\pid$
with $\pid+1$.
The latter ensures that if, concurrently, a new policy (with a different identifier) has been installed, the update
fails and takes no effect.
Algorithm~\ref{alg:simple-update} can therefore also be used
to render our na\"ive load-balancing Algorithm~\ref{alg:naive-lb}
consistent, without any modifications in the update function.

@@

Algorithm~\ref{alg:update} is a more efficient alternative 
to Algorithm~\ref{alg:simple-update}. It uses our id-claimer abstraction
to bound the number of used policy identifiers.
Here, the controller first reads the current policy id and claims to prevent another controller from using it 
for a
different policy (Lines~\ref{update:read-config1}-\ref{update:read-config2}).
%Then it reads the configuration again in case the policy was changed just before claiming its id.
%Then considering other claims the controller chooses
Then the controller claims the first \emph{unused} id and computes the update commands to be executed (Line~\ref{update:compute}).
Finally, within an atomic transaction (\textbf{execatomic}), it verifies whether the chosen policy id has not
been claimed and that
the policy has not been concurrently modified, and then updates the policy.
%[[PK added a note to the id-claimer description
%Note that in Algorithm \ref{alg:update}, the meta-configuration is
%used to store $\textit{pid}$ and id-claimer (the claims).

@@

This paper initiated the study of mechanisms which allow distributed controllers
to synchronize their possibly concurrent actions. We believe that our in-band
approach is natural and attractive as it can avoid the
complexities and limitations of distributed out-of-band agreement protocols.
Moreover, our mechanisms are simple and can be implemented using the \emph{standard OpenFlow}
protocol: they do not require any protocol/hardware extensions as postulated in recent literature~\cite{stn,netpaxos}.
Our work may hence also contribute to the ongoing discussion of what can be implemented
in-band in today's OpenFlow protocol~\cite{reclaim},
as well as to useful high-level concurrency objects and
language abstractions.~\cite{pyretic}
We also plan to release a small library with our synchronization primitives. 
Moreover, we are investigating possible generalizations of our transactional guarantees,
across multiple switches~\cite{stn}, as well as opportunities for a more fine-grained
locking, improving concurrency.

>>>>

\begin{abstract}
Control planes of forthcoming Software-Defined Networks (SDNs)
will be \emph{distributed}: to ensure availability and fault-tolerance,
to improve load-balancing, and to reduce overheads,
modules of the control plane %operating on the data plane 
should be physically distributed.
%a redundant control plane
%can ensure a high availability, and a spatially distributed
%control plane allows to handle time-critical dataplane
%events close to their origins.
%[[PK this is a contraversial point, not important for us
%Moreover, an SDN may be managed by multiple administrators simultaneously.
%]]
However, %a distributed control plane also introduces
%new challenges,
in order to guarantee \emph{consistency} of network operation, actions
performed on the data plane by different controllers may need to be \emph{synchronized},  which is a
 nontrivial task.
%[[PK sounds contradicting to the fact that such systems already exist
%However, not much is known today about how to implement
%such a distributed control plane.
%]]
%
In this paper, we propose a synchronization
framework for control planes  based on the well-known concept of
\emph{transactions}.
%
A transaction aggregates a sequence
of control messages and specific synchronization primitives,
providing all-or-nothing semantics.
We argue that a powerful transactional interface should and can be
implemented \emph{in-band}, using
standard OpenFlow commands on the data plane configuration.
%Indeed, recent OpenFlow versions already come with useful but primitive
%features, in particular \emph{bundling} and \emph{rule-overlap checks}.
%In particular, we present a
%transactional interface which supports arbitrary atomic
%transactions
%over the switch configuration space:
%Our transactions
%can include complex notions of conflicts (beyond flow overlap checks),
%as well as computations and multiple controller interactions.
We discuss how to use our synchronization framework to implement
 consistent policy composition and fault-tolerant control-plane applications.
\end{abstract}

@@

Our synchronization mechanisms can be useful in many situations.
For example, it can be used to implement the fundamental
\emph{Compare-and-Set} (CAS) operation. % on the policy identifier.
When executed within a
transaction, a $\cas(\addr,\textit{old},\textit{new})$ operation 
on a memory location $\addr$
checks if the content of $\addr$ is $\textit{old}$ and if so,
replaces it with $\textit{new}$ %and returns $\true$ 
(the CAS \emph{succeeds}), otherwise it 
%returns $\false$ 
causes an abort of the invoking transaction
(the CAS \emph{fails}).
It is straightforward to implement the CAS operation by executing
$\textit{\compare}(\addr, old)$ followed by $\textit{\memwrite}(\addr,
new)$ within the {\execatomic} construct.

@@

CAS can be used to solve the fundamental
problem of fault-tolerant \emph{consensus}~\cite{FLP85}
\hide{%[[PK, feels like we do not really need this level of detail here
 in which the controllers
need to agree on one of their private inputs, despite failures of an
arbitrary number of controllers.
Indeed, each controller simply performs, within a transaction, a CAS operation to replace the default value
in a memory location at the meta-configuration of a data plane switch.
If the transaction commits, the controller considers itself
the \emph{winner} and decides on its input,
if it aborts (the controller fails in the CAS), it remains to read the
memory to get the input of the winner and decide on it.
}
and allows us
to implement any generic replicated state machine~\cite{Her91}
in the control plane,  in the \emph{wait-free} manner, i.e., tolerating
asynchrony and failure of any number of controllers.
As such, CAS can also be used to implement STN~\cite{stn},
by providing the postulated \emph{read-modify-write} object.

@@

Our mechanism can also be used as a generic template to
implement concurrent policy compositions of a user-specific
$\textsc{\ufunc}()$ function~\cite{stn}.
This way, for example, by choosing our load-balancing
function $\textsc{lb\_update}$ (Example~1 in
Section~\ref{sec:motivation}), 
the pathological scenario described in 
Figure~\ref{fig:CAS-example} can be resolved.
% if we put
%the policy update in our transactional $\execatomic$ construct together
% with a \emph{Compare-and-Set} (CAS) operation on the policy identifier.

Let us first  consider a simple policy update protocol that
allows a set of  controllers to concurrently update switch policies, i.e., sets of
%[[PK policies have been defined in Background]]
effective flow entries, under the
condition that these rules can be \emph{composed} with the currently installed
policy.
Algorithm~\ref{alg:simple-update} uses the bundle feature to update
the switch configuration, where the configuration contains both the
policy and a meta-configuration memory address $\paddr$ that holds the \emph{policy identifier (pid)} .
In the algorithm, the controller first reads the currently installed
policy together with its {\pid},  applies the
$\textsc{\ufunc}()$ function to the current policy which results in a
set of update FlowMod commands and then tries to apply them
atomically together with a CAS operation on $\paddr$ replacing $\pid$
with $\pid+1$.
The latter ensures that if, concurrently, a new policy (with a different identifier) has been installed, the update
fails and takes no effect.
Algorithm~\ref{alg:simple-update} can therefore also be used
to render our na\"ive load-balancing Algorithm~\ref{alg:naive-lb}
consistent, without any modifications in the update function.

@@

Algorithm~\ref{alg:update} is a more efficient alternative 
to Algorithm~\ref{alg:simple-update}. It uses our id-claimer abstraction
to bound the number of used policy identifiers.
Here, the controller first reads the current policy id and claims to prevent another controller from using it 
for a
different policy (Lines~\ref{update:read-config1}-\ref{update:read-config2}).
%Then it reads the configuration again in case the policy was changed just before claiming its id.
%Then considering other claims the controller chooses
Then the controller claims the first \emph{unused} id and computes the update commands to be executed (Line~\ref{update:compute}).
Finally, within an atomic transaction (\textbf{\execatomic}), it verifies whether the chosen policy id has not
been claimed and that
the policy has not been concurrently modified, and then updates the policy.
%[[PK added a note to the id-claimer description
%Note that in Algorithm \ref{alg:update}, the meta-configuration is
%used to store $\textit{pid}$ and id-claimer (the claims).

@@

This paper initiated the study of mechanisms which allow distributed controllers
to synchronize their possibly concurrent actions. We believe that our in-band
approach is natural and attractive as it can avoid the
complexities and limitations of distributed out-of-band agreement protocols.
Moreover, our mechanisms are simple and can be implemented using the \emph{standard OpenFlow}
protocol: they do not require any protocol/hardware extensions as postulated in recent literature~\cite{stn,netpaxos}.
Our work may hence also contribute to the ongoing discussion of what can be implemented
in-band in today's OpenFlow protocol~\cite{reclaim},
as well as to useful high-level concurrency objects and
language abstractions.~\cite{pyretic}
We also plan to release a small library with our synchronization primitives. 
Moreover, we are investigating possible generalizations of our transactional guarantees,
across multiple switches~\cite{stn}, as well as opportunities for more fine-grained
\emph{partial} updates that concern subsets of flow entries, thus improving concurrency.
